Method to infer relatedness using low coverage ancient DNA

### 2022-03-15 paper_overleaf/main.tex

These analyses use cranial and dental metric and morphological and postcranial non-metric data to infer inter-individual relatedness based on the assumption that the phenotype reflects underlying genetic variation and patterns of relatedness. Multifactorial inheritance complicates straightforward interpretations, however, and given the moderate to low (and population specific) heritabilities of most morphometric traits (see Scott & Turner, 1997 for an overview; also Townsend et al., 2009), the effects of development on final phenotypic form are difficult to specify.

@@

Commonly, pairs of related individuals are  identified by looKINg for parts of the genome that are identical by descent (IBD), ie. inherited from a recent common ancestor. Due to the laws of Mendelian segregation, a parent, for example, will share exactly one set of chromosome IBD with its offspring, with the other set of chromosomes in the offspring coming from the other parent. A grandparent will, on average share a quarter of its genome with a grand-child, because recombination broke up the chromosomes.

However, it is not possible to directly observe IBD segments, and so most methods infer them by first identifying segments of the genome that are Identical by State (IBS) and using population allele frequencies to calculate the probability of IBD given IBS \cite{vai_KINship_2020}.  One exception is SNPduo, a software that uses IBS direclty to visualize and analyze pairwise relatedness. \cite{roberson_visualization_2009}. Most methods, however, use IBD as an intermediary in inferring relatedness \cite{boehnke_accurate_1997,lynch_estimation_1999,mcpeek_statistical_2000}. For a non-inbred population, there are three IBD states possible: either no chromosomes are shared, only one chromosome is shared, or both chromosomes are shared. The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals. For example, a pair of siblings are expected to have all three possible IBD states with probabilities (0.25,0.5,0.25) as shown in Fig \ref{fig0:schematic}. These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}. One can also use these probabilities to estimate the coefficient of relatedness as $r= k_1/2 + k_2$. The coefficient of relatedness (r) is defined as the proportion of genome IBD for a pair of individuals.

@@

There are numerous methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to estimate the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}. PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from average of observed IBS, using the allele frequencies at each SNP in samples \cite{purcell_plink_2007}. KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}. .

@@

\subsection{Problems with ancient data, and methods that address these issues}
One major issue with applying the above mentioned methods to ancient DNA is that, frequently, the amount of endogenous DNA is very low, maKINg it very difficult to get genotype calls.
Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods. However, these approaches typically still require at least 2x coverage, since genotype likelihoods may not be very informative at lower coverages. Apart from the low amount of endogenous DNA sequences present, ancient DNA analyses often have additional challenges like contamination from modern populations \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in case of captured sequences. In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations. However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}. Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4x$ coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination from modern humans can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because of comparison of endogenous reads to contaminating reads which would give higher differences than comparison of endogenous reads from individuals from ancient population. It is worth mentioning that different DNA libraries may have different proportion of contamination levels, even when they have been sampled from the same specimen. Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from contaminating population(s). In addition, the target populations may have long runs of homozygosity (ROH) due to a small population size, or recent inbreeding. Long ROH causes related individuals to seem genetically closer to each other, while does not affect the genetic distance between unrelated individuals. This effect may cause an increase in false positives. Moreover, ancient DNA is commonly captured with a SNP array to enrich for endogenous sequences. The way that these SNPs are defined (e.g. selecting SNPs variable in certain individuals) becomes relevant, because relatedness methods based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias. KING robust, for example, differentiates between different relatedness by plotting KING‐robust KINship against fraction of sites in IBS0 \cite{manichaikul_robust_2010}. Another method from Rosenberg,2006 \cite{rosenberg_standardized_2006} uses the fraction of sites in all three IBS states to classify related individuals in the HGDP-CEPH Human Genome Diversity Cell Line Pane. These methods may give different results when applied to data with different ascertainment schemes \cite{waples_allele_2019}.

Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ. This software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual. A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to calculate average pairwise genetic distances, which in turn are used to infer relatedness. The limitations are that the software can identify only up to second degree relatives, without differentiating between siblings and parent-child, which may be crucial when maKINg pedigrees.

\subsection{How our method works}
Here, we present KIN (KINship Inference), a Hidden Markov Model (HMM) to estimate relatedness and IBD from low-coverage ancient DNA data. KIN can detect up to \nth{3} degree relatives, and differentiates between siblings and parent-child. KIN is also able to take into account ROH, contamination and ascertainment bias. We validate the performance of KIN using simulations and show that we are able to infer relatedness in real data from two datasets: a group of Neandertals and a group of Bronze age individuals.

>>>>

@@

Commonly, pairs of related individuals are identified by looking for parts of the genome that are identical by descent (IBD), ie. inherited from a recent common ancestor. Due to the laws of Mendelian segregation, a parent, for example, will share exactly one set of chromosome IBD with its offspring, with the other set of chromosomes in the offspring coming from the other parent. A grandparent will, on average share a quarter of its genome with a grand-child, because recombination broke up the chromosomes.

However, it is not possible to directly observe IBD segments, and so most methods infer them by first identifying segments of the genome that are Identical by State (IBS) and using population allele frequencies to calculate the probability of IBD given IBS \cite{vai_KINship_2020}. One exception is SNPduo, a software that uses IBS direclty to visualize and analyze pairwise relatedness. \cite{roberson_visualization_2009}. Most methods, however, use IBD as an intermediary in inferring relatedness \cite{boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}. For a non-inbred population, there are three IBD states possible: either no chromosomes are shared, only one chromosome is shared, or both chromosomes are shared. The genome-wide proportions of these states (usually referred to as $k_0$, $k_1$, $k_2$, so that $k_0+k_1+k_2=1$) can be used to infer the degree and nature of relatedness for a pair of individuals. For example, a pair of siblings are expected to have all three possible IBD states with probabilities (0.25,0.5,0.25) as shown in Fig \ref{fig0:schematic}. These IBD probabilities can directly be used to categorize their relatedness as shown in table \ref{tab:Table 1}. One can also use these probabilities to estimate the coefficient of relatedness as $r= k_1/2 + k_2$. The coefficient of relatedness (r) is defined as the proportion of genome IBD for a pair of individuals.

@@

There are numerous methods that utilize population allele frequencies, phase information, recombination maps, or genotype calls to estimate the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,li_accurate_2014,thornton_estimating_2012}. PLINK, a popular tool to infer relatedness coefficient in diploid genotype data, estimates IBD probabilities from observed IBS, using the allele frequencies at each SNP in samples \cite{purcell_plink_2007}. KING is another widely used software that allows relatedness inference with genotype data in homogeneous populations using population allele frequencies (KING homo), but can also work with structured populations with no good population allele frequencies using heterozygosity estimates from each sample (KING robust) \cite{manichaikul_robust_2010}. .

@@

\subsection{Problems with ancient data, and methods that address these issues}
One major issue with applying the above mentioned methods to ancient DNA is that, frequently, the amount of endogenous DNA is very low, making it very difficult to get genotype calls.
Several methods surmount this problem by using genotype likelihoods \cite{lipatov_maximum_2015,korneliussen_ngsrelate_2015}.In this way, it is possible to account for the uncertainty in genotype calls by summing over all possible genotypes weighted by their genotype likelihoods. However, these approaches typically still require at least 2x coverage, since genotype likelihoods may not be very informative at lower coverages. Apart from the low amount of endogenous DNA sequences present, ancient DNA analyses often have additional challenges like contamination from modern populations \cite{peyregne_authentict_2020}, absence of a good reference panel, and ascertainment bias in case of captured sequences. In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations. However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}. Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4x$ coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination from modern humans can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because of comparison of endogenous reads to contaminating reads which would give higher differences than comparison of endogenous reads from individuals from ancient population. It is worth mentioning that different DNA libraries may have different proportion of contamination levels, even when they have been sampled from the same specimen. Hence, contamination can reduce the power to detect related pairs, especially when the target population is quite diverged from contaminating population(s). In addition, the target populations may have long runs of homozygosity (ROH) due to a small population size, or recent inbreeding. Long ROH causes related individuals to seem genetically closer to each other, while does not affect the genetic distance between unrelated individuals. This effect may cause an increase in false positives. Moreover, ancient DNA is commonly captured with a SNP array to enrich for endogenous sequences. The way that these SNPs are defined (e.g. selecting SNPs variable in certain individuals) becomes relevant, because relatedness methods based on the fraction of sites in different IBS states are highly sensitive to ascertainment bias. KING robust, for example, differentiates between different relatedness by plotting KING‐robust KINship against fraction of sites in IBS0 \cite{manichaikul_robust_2010}. Another method from Rosenberg,2006 \cite{rosenberg_standardized_2006} uses the fraction of sites in all three IBS states to classify related individuals in the HGDP-CEPH Human Genome Diversity Cell Line Pane. These methods may give different results when applied to data with different ascertainment schemes \cite{waples_allele_2019}.

Several limitations with ancient DNA mentioned above, such as low coverage, the absence of population allele frequencies or a reference panel, and ascertainment bias are resolved by a software called READ. This software addresses the problem of unavailability of genotype calls by randomly sampling alleles from each individual. A string of these alleles at each position (called pseudo-haploids) are then compared to other individuals to calculate average pairwise genetic distances, which in turn are used to infer relatedness. The limitations are that the software can identify only up to second degree relatives, without differentiating between siblings and parent-child, which may be crucial when making pedigrees.

\subsection{How our method works}
Here, we present KIN (Kinship INference), a Hidden Markov Model (HMM) to estimate relatedness and IBD from low-coverage ancient DNA data. KIN can detect up to \nth{3} degree relatives, and differentiates between siblings and parent-child. KIN is also able to take into account ROH, contamination and ascertainment bias. We validate the performance of KIN using simulations and show that we are able to infer relatedness in real data from two datasets: a group of Neandertals and a group of Bronze age individuals.
